---
title: "Mspms_Manusript_Analysis"
author: "Charlie Bayne"
date: "`r Sys.Date()`"
output:
  html_document:
        theme: cosmo
        toc: true
        toc_depth: 6
        toc_float:
            collapsed: true
            smooth_scroll: true
        code_folding: hide
        highlight: tango
        df_print: paged
---

```{r setup, include=FALSE}
library(tidyverse)
ggplot2::theme_set(ggplot2::theme_minimal())
knitr::opts_chunk$set(echo = TRUE,warning = FALSE, message = FALSE)
```

# Processing 


```{r}
# Insure reproducible results.
set.seed(2)
library(mspms)
peaks_data = prepare_peaks("../data/proteomics/peaks_protein-peptides-lfq.csv",
                           colData_filepath = "../data/proteomics/colData_limma.csv")


peaks_processed_qf = process_qf(peaks_data)
peaks_tidy = mspms_tidy(peaks_processed_qf)
```


# Figure 1 - Overview of the mspms package

## Figure 1A - Overview of Mspms package

Generated in plot.IO


## Figure 1B- Example plots?



# Figure 2 - Overall clustering of data


## Figure 2A - PCA 

```{r,fig.height = 4, fig.width =6}
fig2a = plot_pca(peaks_tidy)+
  viridis::scale_color_viridis(discrete = TRUE)

fig2a

ggsave(fig2a,
       filename = "../figures/Figure2_overall_clustering/A_PCA/fig2a.pdf",
       width = 6,
       height = 4)
```


## Figure 2B - unsupervised hierchical clustering of data 

```{r,fig.height = 40, fig.width = 25}
p1 = plot_heatmap(peaks_tidy)
p1
```

saved as .png using the export to png button. 

# Figure 3

## Figure 3A - Table. Prepared in Word. 

## Figure 3B - Volcano Plots

```{r,fig.height = 11, fig.width =2.5}
peaks_stat = limma_stats(peaks_processed_qf)

log2fc_threshold = 3

fig3b <- peaks_stat %>%
    dplyr::mutate(`log_10_p.adj` = -log10(.data$p.adj),
                  time = as.factor(time)) %>%
    ggplot2::ggplot(ggplot2::aes_string(x = "log2fc", y = "log_10_p.adj",color = "time")) +
    ggplot2::geom_point(size = 0.5) +
    ggplot2::geom_hline(
      yintercept = -log10(0.05),
      linetype = "dashed", color = "red"
    ) +
    ggplot2::geom_vline(
      xintercept = log2fc_threshold,
      linetype = "dashed", color = "red"
    ) +
    ggplot2::geom_vline(
      xintercept = -log2fc_threshold,
      linetype = "dashed", color = "red"
    ) +
    ggplot2::labs(x = "Log2 (Time TX/T0)", y = "-log10(p.adj value)") +
  facet_wrap(~condition,ncol = 1)+
  viridis::scale_color_viridis(discrete = TRUE)+
  theme(legend.position = "none")+
  geom_vline(xintercept = 0)

ggplot2::theme_set(ggplot2::theme_minimal())

fig3b

ggsave(fig3b,filename = "../figures/Figure3_experimental_analysis/B/fig3b.pdf",
       width = 2.5, height = 8)

```


```{r}
# Counting number on request of reviewer 3 to add to text

count_peak = peaks_stat %>% dplyr::filter(p.adj <= 0.05,
                                          log2fc > 3) %>% 
  group_by(condition,time) %>% 
  summarise(n = n())
```

## Figure 3C - Cleavage Positions 

```{r,fig.height = 11, fig.width= 2.5}
sig = peaks_stat %>% 
  filter(p.adj <= 0.05, log2fc >3)

d = plot_cleavages_per_pos(sig) 

fig3c =  plot_cleavages_per_pos(sig,ncol = 1 )+
  viridis::scale_color_viridis(discrete = TRUE)+
  theme(legend.position = "none")

fig3c

ggsave("../figures/Figure3_experimental_analysis/C/fig3c.pdf",fig3c, width = 2.5, height = 8)

```


## Figure 3D - Icelogos 

```{r,fig.height = 11,fig.width = 3}
fig3d = plot_all_icelogos(sig)

fig3d

ggsave("../figures/Figure3_experimental_analysis/D/fig3d.pdf",fig3d, 
       width = 3,
       height = 9)


```

# Supplementary Figures 

## Figure S1- QC checks

### A - Full length peptide library detected at T0

```{r,fig.height = 5.5, fig.width = 8}

S1a = plot_qc_check(peaks_processed_qf) +
  theme(legend.position = "bottom")

ggsave("../supplemental/S1_qc_checks/S1a.pdf",S1a,width = 8,
       height = 5.5)

S1a
```


```{r,fig.height = 5.5, fig.width = 8}
nd_peptides = plot_nd_peptides(peaks_processed_qf)

S1b = plot_nd_peptides(peaks_processed_qf)

S1b

ggsave("../supplemental/S1_qc_checks/S1b.pdf",S1b,width = 8,
       height = 5.5)
```

## Figure S2 - Number of signficant peptides per incubation time.

```{r,fig.height =4, fig.width = 5}

sum = sig %>% 
  group_by(condition,time) %>% 
  summarise(n = n()) %>% 
  mutate(condition = as.factor(condition))



S5 = sum %>% 
  ggplot(aes(time,n,color = condition))+
  geom_point()+
  geom_line(aes(group = condition))+
  viridis::scale_color_viridis(discrete = TRUE)+
  theme(legend.position = "bottom")+
  xlab("Time (minutes)")+
  ylab("Number of Peptides")

S5

ggsave("../supplemental/S2_sig_peptides_over_time/S2.pdf",S5,width = 5, height = 4)


```

## Figure S3 - Proteome Software Comparisons 

```{r}
library(mspms)
################################################################################
# Load Data
################################################################################

# Peaks 
## Loaded in previous portion of document. 


# Proteome Dicoverer 
pd_processed = prepare_pd("../data/proteomics/pd_PeptideGroups.txt",
                   colData_filepath = "../data/proteomics/pd_colData.csv") %>% 
  process_qf()


# Fragpipe 
fp_processed = prepare_fragpipe("../data/proteomics/fragpipe_sub_analysis/combined_peptide_mbr_3.tsv",
                               "../data/proteomics/colData_limma.csv") %>% 
  process_qf()



#Sage 

sage_processed = mspms::prepare_sage("../data/proteomics/sage_lfq.tsv",
                                     "../data/proteomics/colData_limma.csv") %>% 
  process_qf()

```


### A- Correlations

```{r,fig.height = 4, fig.width = 8}
################################################################################
# Correlate Values at the Peptide Level — Only include peptides detected in all
################################################################################

# PEAKS
peaks_tidy <- mspms_tidy(peaks_processed_qf,"peptides_log") |>
  filter(!is.na(peptides_log)) %>% 
  group_by(peptide,group,condition,time) %>% 
  summarise(peaks_values = mean(peptides_log))

# Proteome Discoverer (name conversion)
nc <- readxl::read_excel("../data/proteomics/pd_coldata_name_conversion.xlsx")

pd_tidy <- mspms_tidy(pd_processed,"peptides_log") |>
  filter(!is.na(peptides_log)) %>% 
  inner_join(nc, by = c("quantCols" = "pd_quantCols")) |>
  group_by(peptide,group,condition,time) %>% 
  summarise(pd_values = mean(peptides_log))

# FragPipe
fp_tidy <- mspms_tidy(fp_processed,"peptides_log") |>
  filter(!is.na(peptides_log)) %>% 
  group_by(peptide,group,condition,time) %>% 
  summarise(fp_values = mean(peptides_log))

# Sage
sage_tidy <- mspms_tidy(sage_processed,"peptides_log") |>
  filter(!is.na(peptides_log)) %>% 
  group_by(peptide,group,condition,time) %>% 
  summarise(sage_values = mean(peptides_log))

# Combine datasets (intersection only)
all_tidy <- peaks_tidy |>
  inner_join(pd_tidy, by = c("peptide", "condition", "group", "time")) |>
  inner_join(fp_tidy, by = c("peptide","condition", "group", "time")) |>
  inner_join(sage_tidy, by = c("peptide", "condition", "group", "time"))

# Optionally replace 0s with NA (if they represent missing signal)
all_tidy <- all_tidy |>
  mutate(across(ends_with("_values"), ~na_if(., 0))) %>% 
  mutate(peptide_length = as.factor(nchar(gsub("(^.|.$)_?", "", peptide))))
################################################################################
# Correlations
################################################################################

# Common plotting theme
theme_common <- theme_minimal(base_size = 8) +
  theme(plot.title = element_text(hjust = 0.5, size = 10),
        axis.title = element_text(size = 8))

# Helper function for repeated correlations
plot_correlation <- function(df, x, y, label_x, label_y, color = NULL) {
  lm_model <- lm(reformulate(x, y), data = df)
  R2 <- round(summary(lm_model)$r.squared, 2)
  ggplot(df, aes_string(x, y,color = color)) +
    geom_point(size = 0.02, alpha = 0.2) +
    geom_smooth(method = "lm", se = FALSE) +
    geom_abline(slope = 1, linetype = "dashed", color = "red") +
    labs(x = label_x, y = label_y, title = bquote("R"^2*" = "*.(R2))) +
    theme_common
}

################################################################################
# Pairwise comparisons (PEAKS, PD, FP, SAGE)
################################################################################

# PEAKS vs PD
p1 <- plot_correlation(all_tidy, "peaks_values", "pd_values", "PEAKS", "Proteome Discoverer")

# PEAKS vs FragPipe
p2 <- plot_correlation(all_tidy, "peaks_values", "fp_values", "PEAKS", "FragPipe")

# PEAKS vs Sage
p3 <- plot_correlation(all_tidy, "peaks_values", "sage_values", "PEAKS", "Sage")

# Proteome Discoverer vs FragPipe
p4 <- plot_correlation(all_tidy, "pd_values", "fp_values", "Proteome Discoverer", "FragPipe")

# Proteome Discoverer vs Sage
p5 <- plot_correlation(all_tidy, "pd_values", "sage_values", "Proteome Discoverer", "Sage")

# FragPipe vs Sage
p6 <- plot_correlation(all_tidy, "fp_values", "sage_values", "FragPipe", "Sage")


################################################################################
# Arrange and save figure
################################################################################

S3_combined <- ggpubr::ggarrange(
  p1, p2, p3,
  p4, p5, p6,
  ncol = 3, nrow = 2
)

S3_combined

ggsave("../supplemental/S3_proteome_software_comparisons/S3a.pdf",
       S3_combined, height = 2, width = 8)
```

###B - Significant Peptide overlap

```{r,fig.height = 9, fig.width = 5}
################################################################################
# Venn Diagram of Peptides that are significant in each condition
################################################################################

# Run differential stats for all tools
# (peaks stats already computed earlier)
pd_stat <- limma_stats(pd_processed)
fp_stat <- limma_stats(fp_processed)
sage_stat <- limma_stats(sage_processed)


pd_stat <- mspms::limma_stats(pd_processed)
fp_stat <- mspms::limma_stats(fp_processed)
sage_stat <- mspms::limma_stats(sage_processed)
# Extract significant peptides
peaks_sig <- peaks_stat %>%
  filter(p.adj <= 0.05, log2fc > 3) %>%
  mutate(group = paste0(condition, " ", time, " min")) %>%
  mutate(
    peptide_rm = gsub("^._", "", peptide),
    peptide_rm = gsub("_.$", "", peptide_rm),
    peptide_length = nchar(peptide_rm),
    tool = "PEAKS"
  )

pd_sig <- pd_stat %>%
  filter(p.adj <= 0.05, log2fc > 3) %>%
  mutate(group = paste0(condition, " ", time, " min")) %>%
  mutate(
    peptide_rm = gsub("^._", "", peptide),
    peptide_rm = gsub("_.$", "", peptide_rm),
    peptide_length = nchar(peptide_rm),
    tool = "PD"
  )

fp_sig <- fp_stat %>%
  filter(p.adj <= 0.05, log2fc > 3) %>%
  mutate(group = paste0(condition, " ", time, " min")) %>%
  mutate(
    peptide_rm = gsub("^._", "", peptide),
    peptide_rm = gsub("_.$", "", peptide_rm),
    peptide_length = nchar(peptide_rm),
    tool = "FragPipe"
  )

sage_sig <- sage_stat %>%
  filter(p.adj <= 0.05, log2fc > 3) %>%
  mutate(group = paste0(condition, " ", time, " min")) %>%
  mutate(
    peptide_rm = gsub("^._", "", peptide),
    peptide_rm = gsub("_.$", "", peptide_rm),
    peptide_length = nchar(peptide_rm),
    tool = "Sage"
  )

################################################################################
# Generate Venn diagrams per condition
################################################################################

library(ggVennDiagram)

groups_to_plot <- c(
  "cathepsin_A 90 min",
  "cathepsin_B 60 min",
  "cathepsin_C 90 min",
  "cathepsin_D 15 min"
)

plot_list <- list()

for (i in groups_to_plot) {
  
  peaks_f <- peaks_sig %>%
    filter(group == i) %>%
    pull(peptide)
  
  pd_f <- pd_sig %>%
    filter(group == i) %>%
    pull(peptide)
  
  fp_f <- fp_sig %>%
    filter(group == i) %>%
    pull(peptide)
  
  sage_f <- sage_sig %>%
    filter(group == i) %>%
    pull(peptide)
  
  x <- list(
    PEAKS = peaks_f,
    PD = pd_f,
    FragPipe = fp_f,
    Sage = sage_f
  )
  
  p1 <- ggVennDiagram(x, label_size = 3) +
    ggtitle(gsub("[0-9].*", "", i)) +
    scale_fill_distiller(palette = "GnBu") +
    theme(
      legend.position = "none",
      plot.title = element_text(size = 10, hjust = 0.5)
    )
  
  plot_list[[i]] <- p1
}

################################################################################
# Arrange and export
################################################################################

S3b <- ggpubr::ggarrange(
  plot_list[[1]],
  plot_list[[2]],
  plot_list[[3]],
  plot_list[[4]],
  ncol = 1
)

S3b

ggsave(
  "../supplemental/S3_proteome_software_comparisons/S3b.pdf",
  S3b, height = 9, width = 3
) 
```



###C - Cleavage Comparisons


```{r,fig.width = 9, fig.height=6}
peaks_cleavages = plot_cleavages_per_pos(peaks_sig)

S3c1 = peaks_cleavages+
  ggtitle("Peaks Studio")+
  facet_wrap(~condition,ncol = 1,scale = "free_y")+
  theme(legend.position = "none")+
  scale_x_continuous(breaks = 1:13)+
  theme(plot.title = element_text(hjust = 0.5,size = 10))+
  scale_color_viridis_d()

pd_cleavages = plot_cleavages_per_pos(pd_sig)


S3c2 = pd_cleavages+
  ggtitle("Proteome Discoverer")+
  facet_wrap(~condition,ncol = 1,scale = "free_y")+
  theme(legend.position = "none")+
  scale_x_continuous(breaks = 1:14)+
  theme(plot.title = element_text(hjust = 0.5,size = 10))+
    scale_color_viridis_d()


fp_cleavages = plot_cleavages_per_pos(fp_sig)

S3c3 = fp_cleavages+
  ggtitle("Fragpipe")+
  facet_wrap(~condition,ncol = 1,scale = "free_y")+
  theme(legend.position = "none")+
    scale_x_continuous(breaks = 1:14)+
    theme(plot.title = element_text(hjust = 0.5,size = 10),
          axis.title = element_text(size = 10))+
    scale_color_viridis_d()

sage_cleavages = plot_cleavages_per_pos(sage_sig)
S3c4 = sage_cleavages+
  ggtitle("Sage")+
  facet_wrap(~condition,ncol = 1,scale = "free_y")+
  theme(legend.position = "none")+
    scale_x_continuous(breaks = 1:14)+
    theme(plot.title = element_text(hjust = 0.5,size = 10),
          axis.title = element_text(size = 10))+
    scale_color_viridis_d()


S3c = ggpubr::ggarrange(S1c1,S1c2,S1c3,S1c4,ncol = 4)

S3c

ggsave("../supplemental/S3_proteome_software_comparisons/S3c.pdf",S3c,width = 6, height = 9)
```

#S4 Icelogo comparison 

### IceLogo Comparisons

```{r,fig.height = 11,fig.width = 8}

peaks_sig$condition <- factor(
  peaks_sig$condition,
  levels = c("cathepsin_A", "cathepsin_B", "cathepsin_C", "cathepsin_D")
)

peaks_icelogos = plot_all_icelogos(peaks_sig) %>% 
  ggpubr::annotate_figure(top = "Peaks")


pd_sig$condition <- factor(
  pd_sig$condition,
  levels = c("cathepsin_A", "cathepsin_B", "cathepsin_C", "cathepsin_D")
)


pd_icelogos = plot_all_icelogos(pd_sig)%>% 
  ggpubr::annotate_figure(top = "Proteome Discoverer")

fp_sig$condition <- factor(
  fp_sig$condition,
  levels = c("cathepsin_A", "cathepsin_B", "cathepsin_C", "cathepsin_D")
)



fp_icelogos = plot_all_icelogos(fp_sig) %>% 
  ggpubr::annotate_figure(top = "Fragpipe")

sage_sig$condition <- factor(
  sage_sig$condition,
  levels = c("cathepsin_A", "cathepsin_B", "cathepsin_C", "cathepsin_D")
)


sage_icelogos = plot_all_icelogos(sage_sig) %>% 
  ggpubr::annotate_figure(top = "Sage")

S4 = ggpubr::ggarrange(peaks_icelogos,pd_icelogos,fp_icelogos,sage_icelogos,common.legend = TRUE,ncol = 4)

S4

ggsave("../supplemental/S4_proteome_software_icelogos/S4_input.pdf",S2, width =8, height = 11)
```


## Figure S5- Size of Peptides Detected by Each Search Engine

```{r,fig.height = 6, fig.width = 8}

## Significant Peptides
comb = bind_rows(peaks_sig,pd_sig,fp_sig,sage_sig) %>% 
  group_by(peptide_length,condition,tool) %>% 
  summarise(n = n())

p1 = comb %>% 
  ggplot(aes(x = peptide_length,y = n,fill = tool))+
  geom_col()+
  facet_wrap(~condition,scales = "free_y",ncol = 1)+
  scale_x_continuous(breaks = 1:14)+
  ggtitle("Significant Peptides")+
  theme(legend.position = "bottom",
        plot.title = element_text(hjust = 0.5))

p1

peaks_sum_l = mspms_tidy(peaks_processed_qf,"peptides") %>% 
  filter(!is.na(peptides)) %>% 
  mutate(software = "Peaks")


pd_sum_l = mspms_tidy(pd_processed,"peptides") %>% 
  filter(!is.na(peptides)) %>% 
  mutate(software = "PD")


fp_sum_l = mspms_tidy(fp_processed,"peptides") %>% 
    filter(!is.na(peptides)) %>% 
  mutate(software = "Fragpipe")


sage_sum_l = mspms_tidy(sage_processed,"peptides") %>% 
    filter(!is.na(peptides)) %>% 
  mutate(software = "Sage")

comb_2 = bind_rows(peaks_sum_l,pd_sum_l,fp_sum_l,sage_sum_l) %>% 
  group_by(peptide_length,condition,software) %>% 
  summarise(n = n())

p2 = comb_2 %>% 
  ggplot(aes(x = peptide_length,y = n,fill = software))+
  geom_col()+
  facet_wrap(~condition,scales = "free_y",ncol = 1)+
  scale_x_continuous(breaks = 1:14)+
  ggtitle("Detected Peptides")+
  theme(legend.position = "bottom",
        plot.title = element_text(hjust = 0.5))


S5 = ggpubr::ggarrange(p1,p2,common.legend = TRUE,legend = "bottom")

S5

ggsave("../supplemental/S5_proteome_software_peptide_lengths/S5.pdf",S3, width = 8, height = 6)
```








## Figure S6 - Longer motif analysis 

```{r,,fig.height = 11,fig.width = 6}
# Running mspms with 6 residues to the left and right of the cleavage site reported
peaks_prepared_6 = prepare_peaks(lfq_filepath = "../data/proteomics/peaks_protein-peptides-lfq.csv",
                                   colData = "../data/proteomics/colData_limma.csv",
                                   n_residues = 6)

# Calculating all possible cleavages 6 AAs to the left and right of the cleavage
# site.
cleavage_6 = mspms::calculate_all_cleavages(mspms::peptide_library$library_real_sequence,6)


peaks_processed_6 = process_qf(peaks_prepared_6)

peaks_stats_6 = limma_stats(peaks_processed_6)

peaks_sig_6 = peaks_stats_6 %>% 
  filter(p.adj <= 0.05, 
         log2fc >= 3)


S6 = plot_all_icelogos(peaks_sig_6,
                       background_universe = cleavage_6)

S6


ggsave("../supplemental/S6_longer_motif_analysis/S6.pdf",S6, height = 11,
       width =6)

```


# S7 - mspms shiny interface.

Created by taking screenshots of the mspms shiny app. 

# SF6 - generic mspms analysis of cathepsin A-D data. 

```{r,eval = FALSE}
generate_report(peaks_prepared_data,
                output_file = "SF6-generic_mspms_analysis_of_cathepsin_A-D.html",
                outdir = "../supplemental/files/")
```


