---
title: "Analysis"
author: "Charlie Bayne"
date: "`r Sys.Date()`"
output:
  html_document:
        theme: cosmo
        toc: true
        toc_depth: 6
        toc_float:
            collapsed: true
            smooth_scroll: true
        code_folding: hide
        highlight: tango
        df_print: paged
---

```{r setup, include=FALSE}
library(tidyverse)
ggplot2::theme_set(ggplot2::theme_bw())
knitr::opts_chunk$set(echo = TRUE,warning = FALSE, message = FALSE)
```

# Processing 


```{r}
library(mspms)
peaks_data = prepare_peaks("../data/proteomics/peaks_protein-peptides-lfq.csv",
                           colData_filepath = "../data/proteomics/colData.csv")


peaks_processed_qf = process_qf(peaks_data)
peaks_tidy = mspms_tidy(peaks_processed_qf)
```


# Figure 1 - Overview of the mspms package

## Figure 1A - Overview of Mspms package

Generated in plot.IO


## Figure 1B- Example plots?



# Figure 2 - Overall clustering of data


## Figure 2A - PCA 

```{r,fig.height = 4, fig.width =6}
fig2a = plot_pca(peaks_tidy)+
  viridis::scale_color_viridis(discrete = TRUE)

fig2a

ggsave(fig2a,
       filename = "../figures/Figure2_overall_clustering/A_PCA/fig2a.pdf",
       width = 6,
       height = 4)
```


## Figure 2B - unsupervised hierchical clustering of data 

```{r,fig.height = 40, fig.width = 25}
p1 = plot_heatmap(peaks_tidy)
p1
```

saved as .png using the export to png button. 

# Figure 3

## Figure 3A - Volcano Plots

```{r,fig.height = 11, fig.width =2.5}
peaks_t_tests = log2fc_t_test(peaks_processed_qf)

log2fc_threshold = 3

fig3a <- peaks_t_tests %>%
    dplyr::mutate(`log_10_p.adj` = -log10(.data$p.adj),
                  time = as.factor(time)) %>%
    ggplot2::ggplot(ggplot2::aes_string(x = "log2fc", y = "log_10_p.adj",color = "time")) +
    ggplot2::geom_point(size = 0.5) +
    ggplot2::geom_hline(
      yintercept = -log10(0.05),
      linetype = "dashed", color = "red"
    ) +
    ggplot2::geom_vline(
      xintercept = log2fc_threshold,
      linetype = "dashed", color = "red"
    ) +
    ggplot2::geom_vline(
      xintercept = -log2fc_threshold,
      linetype = "dashed", color = "red"
    ) +
    ggplot2::labs(x = "Log2 (Time TX/T0)", y = "-log10(p.adj value)") +
  facet_wrap(~condition,ncol = 1)+
  viridis::scale_color_viridis(discrete = TRUE)+
  theme(legend.position = "bottom")+
  geom_vline(xintercept = 0)

ggplot2::theme_set(ggplot2::theme_minimal())

fig3a

ggsave(fig3a,filename = "../figures/Figure3_experimental_analysis/A/fig3a.pdf",width = 2.5, height = 11)

```

## Figure 3B - Cleavage Positions 

```{r,fig.height = 11, fig.width= 2.5}
sig = peaks_t_tests %>% 
  filter(p.adj <= 0.05, log2fc >3)

d = plot_cleavages_per_pos(sig)

fig3b =  plot_cleavages_per_pos(sig,ncol = 1 )+
  viridis::scale_color_viridis(discrete = TRUE)+
  theme(legend.position = "bottom")

fig3b

ggsave("../figures/Figure3_experimental_analysis/B/fig3b.pdf",fig3b, width = 2.5, height = 11)

```



## Figure 3C - Icelogos 

```{r,fig.height = 11,fig.width = 3.5}
fig3c = plot_all_icelogos(sig)

fig3c

ggsave("../figures/Figure3_experimental_analysis/C/fig3c.pdf",fig3c, 
       width = 3.5,
       height = 11)


```

# Supplementary Figures 

## Figure S1 - Proteome Software Comparisons 



```{r}
library(mspms)
################################################################################
# Load Data
################################################################################

# Peaks 
peaks_processed = prepare_peaks(lfq_filepath = "../data/proteomics/peaks_protein-peptides-lfq.csv",
              colData_filepath = "../data/proteomics/colData.csv") %>% 
  process_qf()



# Proteome Dicoverer 
pd_processed = prepare_pd("../data/proteomics/pd_PeptideGroups.txt",
                   colData_filepath = "../data/proteomics/pd_colData.csv") %>% 
  process_qf()


# Fragpipe 
fp_processed = prepare_fragpipe("../data/proteomics/fragpipe_combined_peptide.tsv",
                               "../data/proteomics/colData.csv") %>% 
  process_qf()


```


### A- Correlations

```{r}
################################################################################
# Correlate Values at the Peptide Level. Only include peptides detected in both
################################################################################


peaks_tidy = mspms_tidy(peaks_processed) %>% 
  rename(peaks_values = peptides_norm) %>% 
  select(peptide,quantCols, peaks_values,condition,group,time)


# need to convert the pd file names to the consistent names as others. 

nc = readxl::read_excel("../data/proteomics/pd_coldata_name_conversion.xlsx")

pd_tidy = mspms_tidy(pd_processed) %>% 
  inner_join(nc, by = c("quantCols" = "pd_quantCols")) %>% 
  select(-quantCols) %>% 
  rename(quantCols = quantCols.y) %>% 
  rename(pd_values = peptides_norm) %>% 
  select(peptide,quantCols,condition,group,time,pd_values)

fp_tidy = mspms_tidy(fp_processed) %>% 
   rename(fp_values = peptides_norm) %>% 
   select(peptide,quantCols,condition,group,time,fp_values)


all_tidy = inner_join(peaks_tidy,pd_tidy,fp_tidy,by = c("peptide","quantCols","condition","group","time"))

all_tidy = inner_join(all_tidy,fp_tidy,by = c("peptide","quantCols","condition","group","time"))


# Peaks vs PD 

lm = lm(peaks_values~pd_values,data = all_tidy)

R2 = round(summary(lm)$r.squared,2)


p1 = all_tidy %>% 
  ggplot(aes(peaks_values,pd_values))+
  geom_point(size = 0.02,alpha = 0.2)+
  geom_smooth(method = "lm")+
  ggtitle("PEAKS vs Proteome Discoverer")+
  annotate("text",label = paste0("R2 = ",R2),x = 100, y = 1000)+
  geom_abline(slope = 1, linetype = "dashed", color = "red")


p1


# Fragpipe vs PEAKS

lm = lm(peaks_values~fp_values,data = all_tidy)

R2 = round(summary(lm)$r.squared,2)


p2 = all_tidy %>% 
  ggplot(aes(peaks_values,pd_values))+
  geom_point(size = 0.02,alpha = 0.2)+
  geom_smooth(method = "lm")+
  ggtitle("PEAKS vs Fragpipe")+
  annotate("text",label = paste0("R2 = ",R2),x = 100, y = 1000)+
  geom_abline(slope = 1, linetype = "dashed", color = "red")


p2

# Proteome Discoverer vs Fragpipe


lm = lm(pd_values~fp_values,data = all_tidy)

R2 = round(summary(lm)$r.squared,2)


p3 = all_tidy %>% 
  ggplot(aes(fp_values,pd_values))+
  geom_point(size = 0.02,alpha = 0.2)+
  geom_smooth(method = "lm")+
  ggtitle("Proteome Discoverer vs Fragpipe")+
  annotate("text",label = paste0("R2 = ",R2),x = 100, y = 1000)+
  geom_abline(slope = 1, linetype = "dashed", color = "red")


p3



```



###B - Significant Peptide overlap

```{r}
################################################################################
#Venn Diagram of Peptides that are signficant in each condition
################################################################################

# Running stats
peaks_stat = log2fc_t_test(peaks_processed)

pd_stat = log2fc_t_test(pd_processed)

fp_stat = log2fc_t_test(fp_processed)
  
  
# Extracting significant peptides 
  
peaks_sig = peaks_stat %>% 
  filter(p.adj <= 0.05, log2fc > 3) %>% 
  mutate(group = paste0(condition,".",time))
  
pd_sig = pd_stat %>% 
    filter(p.adj <= 0.05, log2fc > 3) %>% 
    mutate(group = paste0(condition,".",time))
  
fp_sig  = fp_stat %>% 
    filter(p.adj <= 0.05, log2fc > 3) %>% 
  mutate(group = paste0(condition,".",time))

library(ggVennDiagram)

for(i in unique(pd_sig$group)){
  
  peaks_f = peaks_sig %>% 
    filter(group == i) %>% 
    pull(peptide)
  
  pd_f = pd_sig %>% 
    filter(group == i) %>% 
    pull(peptide)
  
  fp_f = fp_sig %>% 
    filter(group == i) %>% 
    pull(peptide)
  
  
  x <- list(peaks=peaks_f,
          pd=pd_f,
          fp = fp_f)
  
 p1 =  ggVennDiagram(x)+
    ggtitle(i)

  print(p1)
  
}

```


###C - Cleavage Comparisons


```{r}
peaks_cleavages = plot_cleavages_per_pos(peaks_sig)
peaks_cleavages+
  ggtitle("peaks")

pd_cleavages = plot_cleavages_per_pos(pd_sig)
pd_cleavages+
  ggtitle("PD")

fp_cleavages = plot_cleavages_per_pos(fp_sig)
fp_cleavages+
  ggtitle("Fragpipe")
```


### D- IceLogo Comparisons

```{r,fig.height = 11,fig.width = 3.5}
peaks_icelogos = plot_all_icelogos(peaks_sig)
peaks_icelogos

pd_icelogos = plot_all_icelogos(pd_sig)
pd_icelogos

fp_icelogos = plot_all_icelogos(fp_sig)
fp_icelogos

```


## Figure S2 - QC checks

### A - Full length peptide library detected at T0

```{r,fig.height = 5.5, fig.width = 4}

S1a = plot_qc_check(peaks_processed_qf) +
  theme(legend.position = "bottom")

ggsave("../figures/Supplementary/S1_qc_checks/S1a.pdf",S1a,width = 4,
       height = 5.5)

S1a
```


```{r,fig.height = 5.5, fig.width = 8}
nd_peptides = plot_nd_peptides(peaks_processed_qf)

S1c = plot_nd_peptides(peaks_processed_qf)

S1c

ggsave("../figures/Supplementary/S1_qc_checks/S1c.pdf",S1c,width = 8,
       height = 5.5)
```


## Figure S3 - Number of signficant peptides per incubation time.

```{r,fig.height =4, fig.width = 5}

sum = sig %>% 
  group_by(condition,time) %>% 
  summarise(n = n())

S3 = sum %>% 
  ggplot(aes(time,n,color = condition))+
  geom_point()+
  geom_line(aes(group = condition))+
  viridis::scale_color_viridis(discrete = TRUE)+
  theme(legend.position = "bottom")+
  xlab("Time (minutes)")+
  ylab("Number of Peptides")

S3

ggsave("../figures/Supplementary/S3_sig_peptides_over_time/S3.pdf",S3,width = 5, height = 4)


```

## Figure S4- querying described cat C rules

negative enrichment of proline at P1 or P1’, or lysine and arginine at  N-terminal residues. 

### A  
```{r,fig.height = 4, fig.width = 5}


## Checking to see if cleavages that fit the described catC rules are enriched in the data.
fits_rules = peaks_tidy %>% 
  filter(condition == "CatC") %>% 
  filter(grepl("XX[^R|K|X][^P][^P]...",cleavage_seq)) %>% 
  filter(time %in% c(0,90)) %>% 
  dplyr::group_by(Peptide,time,cleavage_seq) %>% 
  summarise(mean = mean(value,na.rm = T)) %>% 
  mutate(time = as.factor(time)) %>% 
  ungroup()


t.test = rstatix::t_test(mean ~ time,data = fits_rules) %>% 
  rstatix::add_y_position() %>% 
  rstatix::add_significance() %>% 
  mutate(y.position = log10(y.position))


S4a = fits_rules %>% 
  ggplot(aes(x = time,y = mean))+
  geom_violin()+
  geom_point()+
  geom_line(aes(group = Peptide),stat = "summary",fun = mean,alpha = 0.3)+
  ggpubr::stat_pvalue_manual(t.test,label = "p.signif")+
  scale_y_log10()+
  xlab("Time (minutes)")

S4a

```

## B
```{r,fig.height = 4, fig.width = 5}


## Checking to see if cleavages that fit the described catC rules are enriched in the data.
fits_rules = mspms_data %>% 
  filter(condition == "CatC") %>% 
  filter(grepl("XX[R|K].....",cleavage_seq)) %>% 
  filter(time %in% c(0,90)) %>% 
  dplyr::group_by(Peptide,time,cleavage_seq) %>% 
  summarise(mean = mean(value,na.rm = T)) %>% 
  mutate(time = as.factor(time)) %>% 
  ungroup()


t.test = rstatix::t_test(mean ~ time,data = fits_rules) %>% 
  rstatix::add_y_position() %>% 
  rstatix::add_significance() %>% 
  mutate(y.position = log10(y.position))


S4b = fits_rules %>% 
  ggplot(aes(x = time,y = mean))+
  geom_violin()+
  geom_point()+
  geom_line(aes(group = Peptide),stat = "summary",fun = mean,alpha = 0.3)+
  ggpubr::stat_pvalue_manual(t.test,label = "p.signif")+
  scale_y_log10()+
  xlab("Time (minutes)")

S4b

```

## C

```{r,fig.height = 4, fig.width = 5}


## Checking to see if cleavages that fit the described catC rules are enriched in the data.
fits_rules = mspms_data %>% 
  filter(condition == "CatC") %>% 
  filter(grepl("XX.[P]....",cleavage_seq)) %>% 
  filter(time %in% c(0,90)) %>% 
  dplyr::group_by(Peptide,time,cleavage_seq) %>% 
  summarise(mean = mean(value,na.rm = T)) %>% 
  mutate(time = as.factor(time)) %>% 
  ungroup()


t.test = rstatix::t_test(mean ~ time,data = fits_rules) %>% 
  rstatix::add_y_position() %>% 
  rstatix::add_significance() %>% 
  mutate(y.position = log10(y.position))


S4c = fits_rules %>% 
  ggplot(aes(x = time,y = mean))+
  geom_violin()+
  geom_point()+
  geom_line(aes(group = Peptide),stat = "summary",fun = mean,alpha = 0.3)+
  ggpubr::stat_pvalue_manual(t.test,label = "p.signif")+
  scale_y_log10()+
  xlab("Time (minutes)")

S4c

```

## D 
```{r,fig.height = 4, fig.width = 5}


## Checking to see if cleavages that fit the described catC rules are enriched in the data.
fits_rules = mspms_data %>% 
  filter(condition == "CatC") %>% 
  filter(grepl("XX..[P]...",cleavage_seq)) %>% 
  filter(time %in% c(0,90)) %>% 
  dplyr::group_by(Peptide,time,cleavage_seq) %>% 
  summarise(mean = mean(value,na.rm = T)) %>% 
  mutate(time = as.factor(time)) %>% 
  ungroup()


t.test = rstatix::t_test(mean ~ time,data = fits_rules) %>% 
  rstatix::add_y_position() %>% 
  rstatix::add_significance() %>% 
  mutate(y.position = log10(y.position))


S4d  = fits_rules %>% 
  ggplot(aes(x = time,y = mean))+
  geom_violin()+
  geom_point()+
  geom_line(aes(group = Peptide),stat = "summary",fun = mean,alpha = 0.3)+
  ggpubr::stat_pvalue_manual(t.test,label = "p.signif")+
  scale_y_log10()+
  xlab("Time (minutes)")

S4d

```

### E

```{r,fig.height = 4, fig.width = 5}
## Reversing the rules in reverse to see if we see enrichment (we expect not to.) ##

fits_rules = mspms_data %>% 
  filter(condition == "CatC") %>% 
  filter(grepl("...[^P][^P][^R|K|X]XX",cleavage_seq)) %>% 
  filter(time %in% c(0,90)) %>% 
  dplyr::group_by(Peptide,time) %>% 
  summarise(mean = mean(value,na.rm = T)) %>% 
  mutate(time = as.factor(time)) %>% 
  ungroup()


t.test = rstatix::t_test(mean ~ time,data = fits_rules) %>% 
  rstatix::add_y_position() %>% 
  rstatix::add_significance() %>% 
  mutate(y.position = log10(y.position))


S4e = fits_rules %>% 
  ggplot(aes(x = time,y = mean))+
  geom_violin()+
  geom_point()+
  geom_line(aes(group = Peptide),stat = "summary",fun = mean,alpha = 0.3)+
  ggpubr::stat_pvalue_manual(t.test,label = "p.signif")+
  scale_y_log10()+
  xlab("Time (minutes)")

S4e

```

```{r,fig.height = 4, fig.width = 5}
## Reversing the rules in reverse to see if we see enrichment (we expect not to.) ##

fits_rules = mspms_data %>% 
  filter(condition == "CatC") %>% 
  filter(!is.na(cleavage_seq)) %>% 
  filter(time %in% c(0,90)) %>% 
  dplyr::group_by(Peptide,time) %>% 
  summarise(mean = mean(value,na.rm = T)) %>% 
  mutate(time = as.factor(time)) %>% 
  ungroup()


t.test = rstatix::t_test(mean ~ time,data = fits_rules) %>% 
  rstatix::add_y_position() %>% 
  rstatix::add_significance() %>% 
  mutate(y.position = log10(y.position))


S4f = fits_rules %>% 
  ggplot(aes(x = time,y = mean))+
  geom_violin()+
  geom_point()+
  geom_line(aes(group = Peptide),stat = "summary",fun = mean,alpha = 0.3)+
  ggpubr::stat_pvalue_manual(t.test,label = "p.signif")+
  scale_y_log10()+
  xlab("Time (minutes)")

S4f

```




```{r,fig.height = 6, fig.width = 10}

S4 = ggpubr::ggarrange(S4a, S4b, S4c, S4d, S4e, S4f, labels = c("A","B","C","D","E","F"))

S4

ggsave("../figures/Supplementary/S4_catC_additional_analysis/S4.pdf",S4, height = 6, width = 10)
```


## Figure S5 - Longer motif analysis 

```{r,,fig.height = 11,fig.width = 6}
# Running mspms with 6 residues to the left and right of the cleavage site reported
peaks_prepared_6 = prepare_peaks(lfq_filepath = "../data/proteomics/peaks_protein-peptides-lfq.csv",
                                   colData = "../data/proteomics/colData.csv",
                                   n_residues = 6)

# Calculating all possible cleavages 6 AAs to the left and right of the cleavage
# site.
cleavage_6 = mspms::calculate_all_cleavages(mspms::peptide_library$library_real_sequence,6)


peaks_processed_6 = process_qf(peaks_prepared_6)

peaks_ttest_6 = log2fc_t_test(peaks_processed_6)

peaks_sig_6 = peaks_ttest_6 %>% 
  filter(p.adj <= 0.05, 
         log2fc >= 3)


S5 = plot_all_icelogos(peaks_sig_6,
                       background_universe = cleavage_6)

S5


ggsave("../figures/Supplementary/S5_longer_motif_analysis/S5.pdf",S5, height = 11,
       width =6)

```


# New Feature

```{r}
# plot_length of peptides 

library(mspms)

peaks_processed = prepare_peaks(lfq_filepath = "../data/proteomics/peaks_protein-peptides-lfq.csv",
              colData_filepath = "../data/proteomics/colData.csv") %>% 
  process_qf()


pd_processed = prepare_pd(peptide_groups_filepath = "../data/proteomics/pd_PeptideGroups.txt",
              colData_filepath = "../data/proteomics/pd_colData.csv") %>% 
  process_qf()

fp_processed = prepare_fragpipe(combined_peptide_filepath = "../data/proteomics/fragpipe_combined_peptide.tsv",
              colData_filepath = "../data/proteomics/colData.csv") %>% 
  process_qf()



prepare_peptide_length_data = function(processed_qf){
  
  out = mspms_tidy(processed_qf, se_name = "peptides") %>% 
    dplyr::filter(!is.na(peptides)) %>% 
    dplyr::mutate(peptide_rm = gsub("^._","",peptide),
                  peptide_rm = gsub("_.$","",peptide_rm),
                  peptide_length = nchar(peptide_rm)) %>% 
    dplyr::select(-peptide_rm)
  
  
  return(out)
  
}


plot_peptide_length = function(processed_qf,plot_type = "intensity"){
  
  plot_data = prepare_peptide_length_data(processed_qf)

  if(plot_type == "intensity"){
    n_rep = plot_data %>% 
   select(quantCols, group) %>% 
   distinct() %>% 
   group_by(group) %>% 
   summarise(n_rep = n())
    
    sum = plot_data %>% 
      inner_join(n_rep,by = c("group"))%>% 
      dplyr::group_by(quantCols,peptide_length,condition,time,n_rep) %>% 
      dplyr::summarize(sum = sum(peptides))
    
    p1 = sum %>% 
      ggplot(aes(peptide_length,sum,color = as.factor(time)))+
      geom_point(position = "dodge")+
      facet_wrap(~condition)+
      scale_y_log10()
    
    return(p1)
    
  } else if(plot_type == "count"){
    
 p2 = plot_data %>% 
   ggplot(aes(peptide_length,fill = as.factor(time)))+
   geom_histogram(position = "dodge")+
   facet_wrap(~condition)
 
 return(p2)
 
  } else{
    stop("plot_type must be intensity or count")
  }
}

plot_peptide_length(peaks_processed,"count")
plot_peptide_length(pd_processed,"count")
plot_peptide_length(fp_processed,"count")


```




